Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-22 paper.tex
updated figure caption
minor fix to spacing

Using these databases, network methods can be applied at various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects \cite{Lounkine:2012yf} others have considered the problem
at the genomic level by characterizing gene networks that are
perturbed by small molecules. Campaigns focused on combination
therapies can also benefit from a network representation and network
inference approaches \cite{Huang:2014by,Tang:2013lr} to identify
target combinations whose simultaneous inhibition leads to synergistic
effects. Mechanism of action studies can benefit from network analyses
by integrating multiple data types. An example of this was described
by Young et al. \cite{Young:2008aa} that integrated high content
screening and ligand-target prediction models to construct network
representations of MoA hypotheses. Network approaches have also been
proposed for SAR analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at
the lead optimization stage. Drug repositioning efforts
\cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network analyses
such as community detection algorithms \cite{girvan2002community}.

@@

Networks constructed from drugs and their protein targets were first
described by Yildirim et al. \cite{Yildirim:2007aa}, who considered
approved and investigational drugs. Networks built using the former
subset highlighted a number of ``follow on'' drugs (i.e., drugs
targeting a protein that is already targeted by another drug. When
investigational drugs were included in the analysis, the resultant
network highlighted a more functionally diverse set of protein
targets. Importantly, the drug-target network representation
effectively highlights the polypharmacology exhibited by many
drugs. They also constructed a drug-target protein network by defining
protein targets as nodes and connecting two nodes if they are targeted
by the same drug. But integrating this network with a protein-protein
interaction network they were able to highlight the fact that drug
target proteins were not as essential \cite{Jeong:2001gd} as high
degree PPI nodes. While many drug-target networks have been
constructed from experimentally derived or manually curated data, a
number of methods to predicting drug-target interactions have also
been described
\cite{Alaimo:2013ev,Wang:2013mw,Heiskanen:2013rc}. However, it is
still an open question as to whether such predicted networks can serve
as the foundation for downstream analyses, though Pahikkala et al. 
\cite{Pahikkala:2014gt} describe a number of guidelines to enhance the
reliability of predicted drug-target interactions.

Many workers have described analytic approaches based on drug-target
networks. For example, the MetaCore and MetaDrug suite of tools have
been used to perform a variety of functional analyses on small
molecules, particularly metabolites \cite{Brennan:2009kx}.  Examples
include pathway enrichment analyses and pathway reconstruction. The
latter is useful in placing new or predicted metabolites in the
context of known protein-protein and protein-compound
interactions. Iwata et al. \cite{Iwata:2013lk} described the use of a
drug-target network to identify protein domains that are associated
with drug side effects. Wang et al. \cite{Wang:2013gn} characterized
the topological properties of drug-target networks and noted that
target essentiality and centrality were strongly linked to drug side
effects and that a drugs' side effects were primarily influenced by
the number of \emph{essential} targets rather than the number of drug
targets. Drug-target networks have been used as input to a variety of
predictive models. For example, a model to predict ATC codes from a
drug-target network was described by Wang et al. 
\cite{Wang:2013ye}. Tabei et al. \cite{Tabei:2012tn} used a drug-target
network as input to a classifier that was used to extract chemogenomic
features. These features represent associations between chemical
substructures and protein domains. The set of features that provides a
compressed representation of the drug-target network and the authors
proposed that they could be used to suggest new substructures for a
given protein domain.

A number of workers have used drug-target network information as the
basis of drug repositioning methodologies. Wang et al.
\cite{Wang:2014lp} who combined drug-target and disease-target
networks to generate repositioning hypotheses. Tan et al. 
\cite{Tan:2014qb} employed drug-target networks as a component of a
larger workflow that included chemical structure similarity, gene
semantic similarity to identify new indications for a number of
preexisting drugs.  Peng et al. \cite{Peng:2014vy} considered a
variant of the drug-target network concept by developing an
interaction network of FDA approved drugs and protein pockets. Drugs
were docked into each pocket and the docking score was used to define
edges between the drug and protein whose pocket it was docked to.

@@

Schadt et al. \cite{Schadt:2009zl} have stressed the importance of
viewing diseases as manifestations of perturbed biological networks
and consequently, the importance of drug discovery approaches that
address the networks underlying a disease state.. Network approaches
to exploring diseases have traditionally explored biological networks
(e.g., protein-protein interaction or gene regulatory networks) in the
context of individual diseases \cite{Lim:2006fe,
  Altieri:2008aa}. Other approaches on the other hand have focused on
classes of disease \cite{Zhang:2013tx} or even all known human
diseases \cite{Goh:2007aa} (the so called ``diseaseome''). While most
of these studies have primarily focused on identifying relationships
between diseases in different ways such as sharing implicated genes
\cite{Bauer-Mehren:2011tv,Goh:2007aa}, sharing symptoms
\cite{Zhou:2014qp} or based on bibliometric associations
\cite{Zhang:2014sh}. Studies on disease networks tend to construct
heterogenous (or $k$-partite) networks where nodes can represent
different classes of entities. An example is the network constructed
from long non-coding RNAs and diseases by Zhou et al. 
\cite{Zhou:2015if}. Based on this network, a random walk approach was
employed to identify novel lncRNA-disease associations. These
approaches play an important role in drug discovery as they provide
opportunities for repurposing drugs based on novel disease-target or
disease-pathway associations. For example, Pacini et al. 
\cite{Pacini:2013kl} describe DvD, a pipeline comprised of R and
Cytoscape \cite{Shannon:2003aa} to identify repurposing opportunities
using gene expression profiles. Another view provided by such analysis
is the druggability of a disease; that is, based on underlying targets
and pathways, is it even feasible to treat the disease using a small
molecule intervention \cite{Boran:2010fk}. An important function of
network analyses of diseases is to go beyond simply identifying new
targets for diseases or links between diseases, but instead attempt to
elucidate the mechanisms underlying condition. As an example of this
is the paradoxical effects of MEK inhibition and PI3K/AKT inhibition
and the effects on cancer resistance \cite{Sos:2009gc}. Together with
knowledge of the underlying network connectivity, these approaches
allow one to design combination treatments - multiple drugs that
affect multiple targets and pathways. This approach was exemplified by
Lehar et al. \cite{Lehar:2009gu} who proposed that drug combination
response surfaces were functionally related to the underlying
connectivity of the drug targets.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{img/salinet} 
  \caption{A SALI network constructed from XXX. See Ref.~\cite{Guha:2008aa} for further details.}
  \label{fig:salinet}
\end{figure}

Network similarity graphs (NSG's) are an alternative network
visualization of SAR data described by Wawer et al. \cite{Wawer:2008aa}
that employs a different characterization of activity cliffs. This
approach is designed to multiplex information by combining
connectivity (representing similarity relationships) with node color
(potency) and size (discontinuity score \cite{Peltason:2007aa}). This
approach has been implemented in the SARANEA tool
\cite{Lounkine:2010fk} and extended \cite{Iyer:2011ij} the NSG concept
to include mechanism of action information.

Networks have been applied to a number of other SAR related
problems. For example, Webb et al. \cite{Webb:2014tp} proposed a
network approach to the interpretation of arbitrary machine learning
models. Hanser et al. \cite{Hanser:2014gl} constructed a network model
to represent SAR knowledge data, resulting in a ``hypothesis
network''. A key feature of this approach is the ability to integrate
disparate data types and build predictive models on top of the
network. Krein and Sukumar \cite{Krein:2011tt} describe an analysis of
chemical spaces using a graph representation and compare chemical
spaces using network metrics.

@@

Most high throughput screening campaigns tend to involve multiple
assays. These usually include a primary screen followed by one or more
secondary screens as well as counter-screens. When data from such
campaigns are deposited into public databases such as Pubchem, the
temporal sequence of the assays is not always maintained. As a result
it can be difficult to track the progression of chemical matter
through the series of assays. In addition, a view of the complete
assay sequence can allow one to detect commonalities between programs
and better compare the performance of screening workflows. Calhoun et
al \cite{Calhoun:2012uq} describe an approach to reconstructing
screening workflows (also termed workflow graphs) from screening
datasets. They employed four heuristic rules to construct graphs from
the data as well as using Pubchem Bioassay meta data and text mining
to identify assays from a given project. These assay networks can be
visualized using an online tool located at
\href{http://swami.wustl.edu/flow}{http://swami.wustl.edu/flow}.
Swamidass et al. \cite{Swamidass:2014vn} describe an approach to
constructing ``assay networks'' based on Pubchem data, where assays
are connected when they exhibit common active (non-promiscuous)
compounds. In this approach connected assays are correlated since they
exhibit compounds with similar activities. This method leads to
connections between apparently unrelated screen allowing one to
identify compounds that exhibit polypharmacological behavior (i.e.,
activity against multiple targets) and thus are candidates for
repurposing. By also labeling screens as phenotypic or biochemical,
this approach allows one to propose target candidates based on
connectivity, thus serving as one approach to target deconvolution.

@@

Varin et al. \cite{Varin:2011ve} describe ``scaffold networks'' which
are similar in nature to scaffold trees but instead of a tree
structure where each scaffold has a single parent, a DAG is
constructed by considering all scaffolds at a given level of the
hierarchy. An example of a scaffold network generated from Alosetron
is shown in Figure \ref{fig:scaffnet}, where the blue structures are
those that would be generated using the Scaffold Tree algorithm and
the green structures would be the extra scaffolds generated using the
Scaffold Network algorithm. Using this approach, together with the
compound set enrichment method \cite{Varin:2010zh} the authors were
able to efficiently identify active scaffold and active molecules

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\linewidth]{img/scaffold-network}
  \caption{An example of the scaffolds generated using
  the Scaffold Tree and Scaffold Network algorithms, starting from
  Alosetron. Blue scaffolds are generated by both methods and green
  scaffolds are generated only by the Scaffold Network
  method. Modified from Figure 1 in Varin et al. \cite{Varin:2011ve}.}
  \label{fig:scaffnet}
\end{figure}

@@

However, a key utility of network approaches is the ability to
integrate different data types (such as small molecules and their
protein targets) in to a single data structure. Drug-target or
drug-disease (or even drug-disease-target) networks are examples of
such approaches. Hartsperger et al. \cite{Hartsperger:2010yg} describe
an approach to fuzzy clustering that is capable of identifying
subgraphs with heterogenous networks. A characteristic feature of such
$k$-partite networks is that it allows one to model the effects of
each class of entity on other classes of entities. This is exemplified
by random walk approaches on bipartite networks such as that described
by Chen et al. \cite{Chen:2012qy} whose proposed a random walk approach
to predict drug-target interactions.

>>>>

Using these databases, network methods can be applied at various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects \cite{Lounkine:2012yf} others have considered the problem
at the genomic level by characterizing gene networks that are
perturbed by small molecules. Campaigns focused on combination
therapies can also benefit from a network representation and network
inference approaches \cite{Huang:2014by,Tang:2013lr} to identify
target combinations whose simultaneous inhibition leads to synergistic
effects. Mechanism of action studies can benefit from network analyses
by integrating multiple data types. An example of this was described
by Young et al.~\cite{Young:2008aa} that integrated high content
screening and ligand-target prediction models to construct network
representations of MoA hypotheses. Network approaches have also been
proposed for SAR analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at
the lead optimization stage. Drug repositioning efforts
\cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network analyses
such as community detection algorithms \cite{girvan2002community}.

@@

Networks constructed from drugs and their protein targets were first
described by Yildirim et al.~\cite{Yildirim:2007aa}, who considered
approved and investigational drugs. Networks built using the former
subset highlighted a number of ``follow on'' drugs (i.e., drugs
targeting a protein that is already targeted by another drug. When
investigational drugs were included in the analysis, the resultant
network highlighted a more functionally diverse set of protein
targets. Importantly, the drug-target network representation
effectively highlights the polypharmacology exhibited by many
drugs. They also constructed a drug-target protein network by defining
protein targets as nodes and connecting two nodes if they are targeted
by the same drug. But integrating this network with a protein-protein
interaction network they were able to highlight the fact that drug
target proteins were not as essential \cite{Jeong:2001gd} as high
degree PPI nodes. While many drug-target networks have been
constructed from experimentally derived or manually curated data, a
number of methods to predicting drug-target interactions have also
been described
\cite{Alaimo:2013ev,Wang:2013mw,Heiskanen:2013rc}. However, it is
still an open question as to whether such predicted networks can serve
as the foundation for downstream analyses, though Pahikkala et al.~
\cite{Pahikkala:2014gt} describe a number of guidelines to enhance the
reliability of predicted drug-target interactions.

Many workers have described analytic approaches based on drug-target
networks. For example, the MetaCore and MetaDrug suite of tools have
been used to perform a variety of functional analyses on small
molecules, particularly metabolites \cite{Brennan:2009kx}.  Examples
include pathway enrichment analyses and pathway reconstruction. The
latter is useful in placing new or predicted metabolites in the
context of known protein-protein and protein-compound
interactions. Iwata et al.~\cite{Iwata:2013lk} described the use of a
drug-target network to identify protein domains that are associated
with drug side effects. Wang et al.~\cite{Wang:2013gn} characterized
the topological properties of drug-target networks and noted that
target essentiality and centrality were strongly linked to drug side
effects and that a drugs' side effects were primarily influenced by
the number of \emph{essential} targets rather than the number of drug
targets. Drug-target networks have been used as input to a variety of
predictive models. For example, a model to predict ATC codes from a
drug-target network was described by Wang et al.~
\cite{Wang:2013ye}. Tabei et al.~\cite{Tabei:2012tn} used a drug-target
network as input to a classifier that was used to extract chemogenomic
features. These features represent associations between chemical
substructures and protein domains. The set of features that provides a
compressed representation of the drug-target network and the authors
proposed that they could be used to suggest new substructures for a
given protein domain.

A number of workers have used drug-target network information as the
basis of drug repositioning methodologies. Wang et al.
\cite{Wang:2014lp} who combined drug-target and disease-target
networks to generate repositioning hypotheses. Tan et al.~
\cite{Tan:2014qb} employed drug-target networks as a component of a
larger workflow that included chemical structure similarity, gene
semantic similarity to identify new indications for a number of
preexisting drugs.  Peng et al.~\cite{Peng:2014vy} considered a
variant of the drug-target network concept by developing an
interaction network of FDA approved drugs and protein pockets. Drugs
were docked into each pocket and the docking score was used to define
edges between the drug and protein whose pocket it was docked to.

@@

Schadt et al.~\cite{Schadt:2009zl} have stressed the importance of
viewing diseases as manifestations of perturbed biological networks
and consequently, the importance of drug discovery approaches that
address the networks underlying a disease state.. Network approaches
to exploring diseases have traditionally explored biological networks
(e.g., protein-protein interaction or gene regulatory networks) in the
context of individual diseases \cite{Lim:2006fe,
  Altieri:2008aa}. Other approaches on the other hand have focused on
classes of disease \cite{Zhang:2013tx} or even all known human
diseases \cite{Goh:2007aa} (the so called ``diseaseome''). While most
of these studies have primarily focused on identifying relationships
between diseases in different ways such as sharing implicated genes
\cite{Bauer-Mehren:2011tv,Goh:2007aa}, sharing symptoms
\cite{Zhou:2014qp} or based on bibliometric associations
\cite{Zhang:2014sh}. Studies on disease networks tend to construct
heterogenous (or $k$-partite) networks where nodes can represent
different classes of entities. An example is the network constructed
from long non-coding RNAs and diseases by Zhou et al.~
\cite{Zhou:2015if}. Based on this network, a random walk approach was
employed to identify novel lncRNA-disease associations. These
approaches play an important role in drug discovery as they provide
opportunities for repurposing drugs based on novel disease-target or
disease-pathway associations. For example, Pacini et al.~
\cite{Pacini:2013kl} describe DvD, a pipeline comprised of R and
Cytoscape \cite{Shannon:2003aa} to identify repurposing opportunities
using gene expression profiles. Another view provided by such analysis
is the druggability of a disease; that is, based on underlying targets
and pathways, is it even feasible to treat the disease using a small
molecule intervention \cite{Boran:2010fk}. An important function of
network analyses of diseases is to go beyond simply identifying new
targets for diseases or links between diseases, but instead attempt to
elucidate the mechanisms underlying condition. As an example of this
is the paradoxical effects of MEK inhibition and PI3K/AKT inhibition
and the effects on cancer resistance \cite{Sos:2009gc}. Together with
knowledge of the underlying network connectivity, these approaches
allow one to design combination treatments - multiple drugs that
affect multiple targets and pathways. This approach was exemplified by
Lehar et al.~\cite{Lehar:2009gu} who proposed that drug combination
response surfaces were functionally related to the underlying
connectivity of the drug targets.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{img/salinet} 
  \caption{A SALI network constructed from set of 62 dihydroquinoline
    derivatives that were designed to inhibit the glucocorticoid
    receptor. See Ref.~\cite{Guha:2008aa} for further details.}
  \label{fig:salinet}
\end{figure}

Network similarity graphs (NSG's) are an alternative network
visualization of SAR data described by Wawer et al.~\cite{Wawer:2008aa}
that employs a different characterization of activity cliffs. This
approach is designed to multiplex information by combining
connectivity (representing similarity relationships) with node color
(potency) and size (discontinuity score \cite{Peltason:2007aa}). This
approach has been implemented in the SARANEA tool
\cite{Lounkine:2010fk} and extended \cite{Iyer:2011ij} the NSG concept
to include mechanism of action information.

Networks have been applied to a number of other SAR related
problems. For example, Webb et al.~\cite{Webb:2014tp} proposed a
network approach to the interpretation of arbitrary machine learning
models. Hanser et al.~\cite{Hanser:2014gl} constructed a network model
to represent SAR knowledge data, resulting in a ``hypothesis
network''. A key feature of this approach is the ability to integrate
disparate data types and build predictive models on top of the
network. Krein and Sukumar \cite{Krein:2011tt} describe an analysis of
chemical spaces using a graph representation and compare chemical
spaces using network metrics.

@@

Most high throughput screening campaigns tend to involve multiple
assays. These usually include a primary screen followed by one or more
secondary screens as well as counter-screens. When data from such
campaigns are deposited into public databases such as Pubchem, the
temporal sequence of the assays is not always maintained. As a result
it can be difficult to track the progression of chemical matter
through the series of assays. In addition, a view of the complete
assay sequence can allow one to detect commonalities between programs
and better compare the performance of screening workflows. Calhoun et
al \cite{Calhoun:2012uq} describe an approach to reconstructing
screening workflows (also termed workflow graphs) from screening
datasets. They employed four heuristic rules to construct graphs from
the data as well as using Pubchem Bioassay meta data and text mining
to identify assays from a given project. These assay networks can be
visualized using an online tool located at
\href{http://swami.wustl.edu/flow}{http://swami.wustl.edu/flow}.
Swamidass et al.~\cite{Swamidass:2014vn} describe an approach to
constructing ``assay networks'' based on Pubchem data, where assays
are connected when they exhibit common active (non-promiscuous)
compounds. In this approach connected assays are correlated since they
exhibit compounds with similar activities. This method leads to
connections between apparently unrelated screen allowing one to
identify compounds that exhibit polypharmacological behavior (i.e.,
activity against multiple targets) and thus are candidates for
repurposing. By also labeling screens as phenotypic or biochemical,
this approach allows one to propose target candidates based on
connectivity, thus serving as one approach to target deconvolution.

@@

Varin et al.~\cite{Varin:2011ve} describe ``scaffold networks'' which
are similar in nature to scaffold trees but instead of a tree
structure where each scaffold has a single parent, a DAG is
constructed by considering all scaffolds at a given level of the
hierarchy. An example of a scaffold network generated from Alosetron
is shown in Figure \ref{fig:scaffnet}, where the blue structures are
those that would be generated using the Scaffold Tree algorithm and
the green structures would be the extra scaffolds generated using the
Scaffold Network algorithm. Using this approach, together with the
compound set enrichment method \cite{Varin:2010zh} the authors were
able to efficiently identify active scaffold and active molecules

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\linewidth]{img/scaffold-network}
  \caption{An example of the scaffolds generated using
  the Scaffold Tree and Scaffold Network algorithms, starting from
  Alosetron. Blue scaffolds are generated by both methods and green
  scaffolds are generated only by the Scaffold Network
  method. Modified from Figure 1 in Varin et al.~\cite{Varin:2011ve}.}
  \label{fig:scaffnet}
\end{figure}

@@

However, a key utility of network approaches is the ability to
integrate different data types (such as small molecules and their
protein targets) in to a single data structure. Drug-target or
drug-disease (or even drug-disease-target) networks are examples of
such approaches. Hartsperger et al.~\cite{Hartsperger:2010yg} describe
an approach to fuzzy clustering that is capable of identifying
subgraphs with heterogenous networks. A characteristic feature of such
$k$-partite networks is that it allows one to model the effects of
each class of entity on other classes of entities. This is exemplified
by random walk approaches on bipartite networks such as that described
by Chen et al.~\cite{Chen:2012qy} whose proposed a random walk approach
to predict drug-target interactions.
